Tropical Mathematics and the Lambda Calculus

### 2023-05-14 section5.tex
Committ
Ho corretto un minierrore nella versione LICS, non si sa mai la usiamo per le future versioni e ci scordavamo di correggerlo

Here we follow a similar patter: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.

@@

The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}. Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$, 
the value of the interpretation $\model{M}\in \Lawv^{\BB N}$ on a particular natural number $k$, i.e.~$\model{M}(k)\in \Lawv$, corresponds to the \emph{minimum} number of $\beta$- or $\TT{fix}$-redexes reduced in a reductions sequence from $M$ to $\underline n$. 
In the next paragraph we will illustrate an analogous ``best case'' analysis for probabilistic programs.

@@

Secondly, the Lipschitz conditions from Section \ref{section4} allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$. 
Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
\begin{align}
\TT C[M] = \sum_{i=0}^{k} \TT D^{(k)}\Big[\lambda x.\TT C[x]\Big]\cdot M^{k}
\end{align}
and similarly for $\TT C[N]$. It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high. 
Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{Ehrhard2005}, 
a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
We mention this point in Section~\ref{section8}

@@

Let us consider in this subsection a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
\[
 \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
 \qquad\qquad
 \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
\]
We add reduction rules:
\[
 M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
\]
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing through $\trop$ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

>>>>

Here we follow a similar pattern: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.

@@

The possibility of using the relational model over the tropical semiring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}. Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$, 
the value of the interpretation $\model{M}\in \Lawv^{\BB N}$ on a particular natural number $k$, i.e.~$\model{M}(k)\in \Lawv$, corresponds to the \emph{minimum} number of $\beta$- or $\TT{fix}$-redexes reduced in a reductions sequence from $M$ to $\underline n$. 
In the next paragraph we will illustrate an analogous ``best case'' analysis for probabilistic programs.

@@

Secondly, the Lipschitz conditions from Section \ref{section4} allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$. 
Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
\begin{align}
\TT C[M] = \sum_{i=0}^{k} \TT D^{(k)}\Big[\lambda x.\TT C[x]\Big]\cdot M^{k}
\end{align}
and similarly for $\TT C[N]$. It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always \emph{finite}, although its non-zero coefficients may be arbitrarily high. 
Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{Ehrhard2005}, 
a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
%We mention this point in Section~\ref{section8}.

@@

Let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
\[
 \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
 \qquad\qquad
 \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
\]
We add reduction rules:
\[
 M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
\]
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing through $\trop $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.
